---
title: "R for Marine Science Workshop 2"
author: "Darius HUISMAN"
date: "2026-06-04"
output: 
 html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 6
    fig_width: 10
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
# Set options, housekeeping
knitr::opts_chunk$set(
	echo = FALSE,
	include = TRUE,
	message = FALSE,
	warning = FALSE)
```

## Introduction
This qmd file covers the following topics for Workshop 2 of the R for Marine Science Module:
1. Tidying data
2. Pivoting data to make it tidy
3. Lengthening datasets
4. Widening datasets
5. Separating and uniting data tables
6. Wrangling strings and dates
7. Relational data: Joining tables
8. Handling missing values
9. Practical exercises: Penguins, pivots, and relational data

## Clear out environment

```{r housekeeping message=FALSE, warning=FALSE, include=FALSE}
rm(list=ls())
```

## Load packages
```{r packages, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(lubridate)
library(palmerpenguins)
library(ggplot2)
```


#2.2 Tidying data

```{r explore table1 table2 and table3}
table1 <- table1
table1

table2 <- table2
table2

table3 <- table3
table3

```

```{r Tidy table1 and visualise data}
# Compute rate per 10,000
table1 %>% 
  mutate(rate = cases / population * 10000)

# Compute cases per year
table1 %>% 
  count(year, wt = cases)

# Visualise changes over time
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))
```

#2.3 Pivoting data to make it tidy

#2.4 Lengthening datasets

```{r pivotlonger()}
table4 <- table4a
table4

pivot_longer(table4, c("1999", "2000"), names_to = "year", values_to = "cases")
```

```{r billboard, pivotlonger()}
billboard <- billboard
billboard
# Pivot to tidy format
billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank"
  )

# Pivot to tidy format, dropping NAs  
billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank",
    values_drop_na = TRUE
  )
```

```{r Pivoting longer example}
# Create a simple data frame
df <- tribble(
  ~id,  ~bp1, ~bp2,
   "A",  100,  120,
   "B",  140,  115,
   "C",  120,  125
)

# Pivot to tidy format
df |> 
  pivot_longer(
    cols = bp1:bp2,
    names_to = "measurement",
    values_to = "value"
  )
```


#2.5 Widening datasets

```{r widening datasets}
cms_patient_experience

# See complete set of values
cms_patient_experience |> 
  distinct(measure_cd, measure_title)

# Pivot to wide format
cms_patient_experience |> 
  pivot_wider(
    names_from = measure_cd,
    values_from = prf_rate
  )

# Pivot to wide format, keeping org variables
cms_patient_experience |> 
  pivot_wider(
    id_cols = starts_with("org"),
    names_from = measure_cd,
    values_from = prf_rate
  )
```

```{r Pivoting wider - example}
# Create a simple data frame
df <- tribble(
  ~id, ~measurement, ~value,
  "A",        "bp1",    100,
  "B",        "bp1",    140,
  "B",        "bp2",    115, 
  "A",        "bp2",    120,
  "A",        "bp3",    105
)

# Pivot to wide format 
df |> 
  pivot_wider(
    names_from = measurement,
    values_from = value
  )

# Pivot to wide format, keeping id variable
df |> 
  distinct(measurement) |> 
  pull()

# Pivot to wide format, keeping id variable and dropping measurement variable
df |> 
  select(-measurement, -value) |> 
  distinct()

# Pivot to wide format, keeping id variable and dropping measurement variable, and adding new variables
df |> 
  select(-measurement, -value) |> 
  distinct() |> 
  mutate(x = NA, y = NA, z = NA)
```

#2.6 Pivoting Exercise using Palmer Penguins

```{r Lengthening data with pivot_longer()}
# Create a long version of the penguins dataset

penguins_long <- penguins |>
  pivot_longer(
    cols = c(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g),
    names_to = "measurement_type",
    values_to = "value"
  )
# View the result
head(penguins_long)

# Visualise the distribution of measurements across species
penguins_long |>
  drop_na(value) |>
  ggplot(aes(x = value, fill = species)) +
  geom_histogram(bins = 30, alpha = 0.7, colour = "black") +
  facet_wrap(~ measurement_type, scales = "free_x") +
  theme_minimal() +
  labs(
    title = "Morphometric distributions across penguin species",
    x = "Measurement value",
    y = "Frequency"
  )
```

```{r Widening data with pivot_wider()}
# Create a wide version of the penguins dataset
mass_summary <- penguins |>
  drop_na(body_mass_g) |>
  group_by(species, island) |>
  summarise(mean_mass = mean(body_mass_g))

head(mass_summary)

# Pivot to wide format to create a matrix of mean body mass by species and island
mass_matrix <- mass_summary |>
  pivot_wider(
    names_from = island,
    values_from = mean_mass
  )

head(mass_matrix)
```


#2.7 Separating and uniting data tables

```{r Separating and uniting data tables}
table3

# Separate the 'rate' column into 'cases' and 'population'
table3 %>% 
  separate(rate, into = c("cases", "population"))

# Separate the 'rate' column into 'cases' and 'population', converting to numeric
table3 %>% 
  separate(rate, into = c("cases", "population"), sep = "/")

# Separate the 'rate' column into 'cases' and 'population', converting to numeric
table3 %>% 
  separate(rate, into = c("cases", "population"), convert = TRUE)

# Separate the 'year' column into 'century' and 'year'
table3 %>% 
  separate(year, into = c("century", "year"), sep = 2)

# Separate the 'year' column into 'century' and 'year', converting to numeric
table5 %>% 
  unite(new, century, year, sep = "")
```


#2.8 Wrangling strings and dates

```{r Standardising text with stringr}
# A remarkably messy data frame of field sites
messy_sites <- tibble(
  site_id = c(" Nelly Bay", "nelly_bay", "NELLY BAY", " Geoffrey_Bay ", "geoffrey bay")
)

# Using stringr within mutate to standardize the text
clean_sites <- messy_sites |>
  mutate(
    # 1. Convert everything to lowercase
    site_clean = str_to_lower(site_id),
    # 2. Trim any leading or trailing whitespace (invisible spaces at the ends)
    site_clean = str_trim(site_clean),
    # 3. Replace any spaces with underscores
    site_clean = str_replace_all(site_clean, pattern = " ", replacement = "_")
  )

print(clean_sites)

```

```{r Mastering time with lubridate}
# Parsing different date formats
date_1 <- dmy("25/12/2026")
date_2 <- ymd("2026-12-25")

# R now recognizes these as identical Date objects
date_1 == date_2

# A tibble of raw sensor data with a messy character timestamp
sensor_data <- tibble(
  raw_time = c("14-05-2026 08:30:00", "14-05-2026 08:45:00", "14-05-2026 09:00:00"),
  temperature = c(24.5, 24.6, 24.4)
)

# Converting character strings to true POSIXct datetime objects
sensor_clean <- sensor_data |>
  mutate(
    true_time = dmy_hms(raw_time)
  )

print(sensor_clean)
```

#2.9 Relational data: Joining tables

```{r Relational data: Joining tables}
# Table 1: Biological observation data
observations <- tibble(
  site_code = c("NB", "GB", "MI", "NB", "HB"),
  species = c("Trout", "Snapper", "Trout", "Cod", "Trout"),
  count = c(5, 2, 1, 3, 8)
)

# Table 2: Spatial metadata
site_metadata <- tibble(
  site_code = c("NB", "GB", "MI", "RP", "WP"),
  zone = c("Marine National Park", "Conservation Park", "Habitat Protection", "General Use", "Other Use"),
  lat = c(-19.16, -19.15, -19.14, -19.12, -19.11)
)
```

```{r The workhorse: left_join()}
# Joining metadata to our observations
joined_data <- observations |>
  left_join(site_metadata, by = join_by(site_code))

print(joined_data)
```

```{r The strict filter: inner_join()}
# Joining metadata to our observations, but only keeping rows with matches in both tables
matched_data <- observations |>
  inner_join(site_metadata, by = join_by(site_code))
glimpse(matched_data)
```

```{r The detective: anti_join()}
# Which observations are missing from our metadata dictionary?
missing_context <- observations |>
  anti_join(site_metadata, by = join_by(site_code))
missing_context
```


#2.10 Handling missing values

```{r Converting legacy errors with na_if()}
# Raw data from an old temperature logger
logger_data <- tibble(
  depth_m = c(10, 20, 30, 40),
  temp_c = c(24.5, 24.1, -999, 23.5)) # -999 is a known sensor error code

# Convert the -999 error codes to true NA values
fixed_logger <- logger_data |>
  mutate(temp_c = na_if(temp_c, -999))

print(fixed_logger)
```

```{r Replacing missing values with coalesce()}
# Simulate some count data
shark_counts <- tibble(
  site = c("Reef_A", "Reef_B", "Reef_C"),
  shark_count = c(3, NA, 5)) # The NA here actually means 0 sharks were seen

# Replace NA with 0 
shark_fixed <- shark_counts |>
  mutate(shark_count = coalesce(shark_count, 0))

print(shark_fixed)
```

```{r The special case of NaN (Not a Number)}
cpue_data <- tibble(
  site = c("Bay_1", "Bay_2"),
  catch = c(10, 0),
  effort_hours = c(2, 0))

# Calculating CPUE (catch / effort)
cpue_calc <- cpue_data |>
  mutate(
    cpue = catch / effort_hours)

print(cpue_calc)
```

```{r Implicit missing values and the zero-data framework: complete()}
raw_catch <- tibble(
  site = c("Reef_1", "Reef_1", "Reef_2"),
  species = c("Pmaculatus", "Pleopardus", "Pmaculatus"),
  count = c(5, 2, 8)) 

# Force inclusion of hidden zero-catch data
full_catch_matrix <- raw_catch |>
  complete(site, species, fill = list(count = 0))

print(full_catch_matrix)
```

```{r The nuclear option: drop_na()}
sensor_log <- tibble(
  day = 1:4,
  salinity = c(35.2, 35.1, NA, 35.3)
)

# Remove any row where salinity is missing
clean_log <- sensor_log |>
  drop_na(salinity) 
print(clean_log)
```


#2.11 Practical Exercises: Penguins, Pivots, and Relational Data

##Exercise 1: Cleaning the messy metadata
Task: Create a new object called clean_metadata. Using mutate(), stringr, and lubridate, fix the island_name column so that it perfectly matches the island column in the penguins dataset (hint: watch out for those hidden spaces!). Then, convert the station_install column into a proper Date object.

```{r Ex1: Cleaning the messy metadata}
# The messy metadata provided by your colleague
island_metadata <- tibble(
  island_name = c(" biscoe", "Dream ", "Torgersen"),
  station_install = c("15/01/2003", "22-03-2004", "05/11/2001"),
  latitude = c(-64.81, -64.73, -64.76)
)

print(island_metadata)

print(penguins)

# Use mutate, stringr, and lubridate to clean the data
clean_metadata <- island_metadata |>
  mutate(
    # 1. Make sure all island names start with a capital letter, and the rest are lowercase
    island_name = str_to_title(island_name),
    # 2. Trim any leading or trailing whitespace (invisible spaces at the ends)
    island_name = str_trim(island_name),
    # 3. Convert station_install to Date forma
    station_install = dmy(station_install)
      )

print(clean_metadata)
```

##Exercise 2: The Relational Join
Task: Create a new object called penguins_spatial. Use a left_join() to attach your clean_metadata to the main penguins dataset. Hint: Because the column containing the island names is called island in the penguins dataset, and island_name in your metadata, you will need to specify how they match using by = join_by(island == island_name).

```{r Ex2: The Relational Join}
penguins_spatial <- penguins |>
  left_join(clean_metadata, by = join_by(island == island_name))

# Scroll to the right to see the new columns
head(penguins_spatial)
```

##Exercise 3: The Wide Summary Matrix
Task: Starting with your penguins_spatial dataset:
Drop any missing values in the body_mass_g column.
  1. Group by species and island.
  2. Summarise the data to find the maximum body mass (max()).
  3. Pivot the dataset wider so that the species form the rows, and the islands form the columns.

```{r Ex3: The Wide Summary Matrix}
# Create a wide summary matrix of maximum body mass by species and island
penguins_summary <- penguins_spatial |>
  # drop any missing values in the body_mass_g column
  drop_na(body_mass_g) |>
  # group by species and island
  group_by(species, island) |>
  # summarise the data to find the maximum body mass (max())
  summarise(max_mass = max(body_mass_g)) |>
  pivot_wider(
    names_from = island,
    values_from = max_mass
  )

print(penguins_summary)
```












