---
title: "R for Marine Science Workshop 2 Exercise 2.12"
author: "Darius HUISMAN"
date: "2026-06-04"
output: 
 html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 6
    fig_width: 10
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
# Set options, housekeeping
knitr::opts_chunk$set(
	echo = FALSE,
	include = TRUE,
	message = FALSE,
	warning = FALSE)
```

# Introduction
This qmd file covers the following point 12 (Keystone exercise: Estuary Fish Survey - Data Rescue Mission) of Workshop 2 of the R for Marine Science Module.

The goal is to write a single, reproducible R script (or .qmd) that ingests this mess, cleans it, joins it into a single master dataset, and produces a publication-ready visualisation of the estuary's ecology.

## Clear out environment

```{r housekeeping message=FALSE, warning=FALSE, include=FALSE}
rm(list=ls())
```

## Load packages
Load the packages necessary to this project
```{r packages, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(lubridate)
library(readxl)
library(ggplot2)
```

# Phase 1: Ingestion and Decontamination

## Load datasets
```{r Load datasets}
catch_site1 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 1)
catch_site1
catch_site2 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 2)
catch_site2
catch_site3 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 3)
catch_site3
catch_site4 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 4)
catch_site4

catch_log <- bind_rows(catch_site1, catch_site2, catch_site3,catch_site4)
catch_log

metadata <- read_csv("../data/workshop2/estuary_metadata.csv")
metadata
sonde_data <- read_csv("../data/workshop2/estuary_sonde_data.csv")
sonde_data
species_dictionary <- read_csv("../data/workshop2/species_dictionary.csv")
species_dictionary
```

### catch_log
```{r catch_log Ingestion and Decontamination}
# Investigate dataset catch_log
catch_log
head(catch_log)
glimpse(catch_log)
str(catch_log)
```

```{r catch_log Ingestion and Decontamination}
# Standardise the site and species names to lowercase with underscores
clean_catch_log <- catch_log |>
  mutate(
    #site names
    # 1. Make sure all site names are in lowercase
    site = str_to_lower(site),
    # 2. Trim any leading or trailing whitespace (invisible spaces at the ends)
    site = str_trim(site),
    # 3. Replace any spaces with underscores
    site = str_replace_all(site, pattern = " ", replacement = "_"),
    #species names
    # 1. Make sure all site names are in lowercase
    species = str_to_lower(species),
    # 2. Trim any leading or trailing whitespace (invisible spaces at the ends)
    species = str_trim(species),
    # 3. Replace any spaces with underscores
    species = str_replace_all(species, pattern = " ", replacement = "_")
      )
print(clean_catch_log)
```

###sonde_data
```{r sonde_data Ingestion and Decontamination}
# Investigate dataset catch_log
sonde_data
head(sonde_data)
glimpse(sonde_data)
str(sonde_data)
```

```{r sonde_data Temporal Parsing & Hardware Error}
# Parse dates into formal Date-Time objects
clean_sonde_data <- sonde_data |>
  mutate(
    # Parse dates into formal Date-Time objects
    timestamp = dmy_hm(timestamp),
    # Convert the -999 error codes to true NA values in the turbidity column
    turbidity = na_if(turbidity, -999)
  )
clean_sonde_data
```


# Phase 2: The Relational Architecture

