---
title: "R for Marine Science Workshop 2 Exercise 2.12"
author: "Darius HUISMAN"
date: "2026-06-04"
output: 
 html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 6
    fig_width: 10
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
# Set options, housekeeping
knitr::opts_chunk$set(
	echo = FALSE,
	include = TRUE,
	message = FALSE,
	warning = FALSE)
```

# Introduction
This qmd file covers the following point 12 (Keystone exercise: Estuary Fish Survey - Data Rescue Mission) of Workshop 2 of the R for Marine Science Module.

The goal is to write a single, reproducible R script (or .qmd) that ingests this mess, cleans it, joins it into a single master dataset, and produces a publication-ready visualisation of the estuary's ecology.

## Clear out environment

```{r housekeeping message=FALSE, warning=FALSE, include=FALSE}
rm(list=ls())
```

## Load packages
Load the packages necessary to this project
```{r packages, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(lubridate)
library(readxl)
library(ggplot2)
```

# Phase 1: Ingestion and Decontamination

## Load datasets
```{r Load datasets}
catch_1 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 1)
catch_1
catch_2 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 2)
catch_2
catch_3 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 3)
catch_3
catch_4 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 4)
catch_4
metadata <- read_csv("../data/workshop2/estuary_metadata.csv")
metadata
sonde_data <- read_csv("../data/workshop2/estuary_sonde_data.csv")
sonde_data
species_dictionary <- read_csv("../data/workshop2/species_dictionary.csv")
species_dictionary
```
Excel Tab Iteration: The catch log is split across multiple tabs (one for each site). You will need to figure out how to read all tabs and bind them together into a single data frame.
```{r Excel Tab Iteration}
# Bind the four catch data sheets (catch1, catch2, catch3 & catch4) together
all_catch <- 
```





