---
title: "R for Marine Science Workshop 4"
author: "Darius HUISMAN"
date: "2026-06-11"
output: 
 html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 6
    fig_width: 10
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
# Set options, housekeeping
knitr::opts_chunk$set(
	echo = FALSE,
	include = TRUE,
	message = FALSE,
	warning = FALSE)
```

# Introduction



## Clear out environment

```{r housekeeping message=FALSE, warning=FALSE, include=FALSE}
rm(list=ls())
```

## Load packages

Load the packages necessary to this project

```{r packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(sf) # simple features
library (terra) # for raster
library(tmap) # Thematic maps are geographical maps in which spatial data distributions are visualized
library(leaflet) # for interactive maps
```

## Loadling data

```{r load data, message=FALSE, warning=FALSE}
data <- read_csv("../data/workshop4/copepods_raw.csv")
data
```

# Data exploration

## Check coordinates

```{r check coordinates}
ggplot(data) + 
  aes(x = longitude, y = latitude, color = richness_raw) +
  geom_point()
```

## Check for other patterns

```{r other patterns}
ggplot(data, aes(x = latitude, y = richness_raw)) + 
  stat_smooth() + 
  geom_point()
```


# Getting going with maps

```{r turn data into sf}
sdata <- st_as_sf(data, coords = c("longitude", "latitude"), 
                 crs = 5126)
```


# Coordinates reference system

```{r coordiantes reference system}
crs5126 <- st_crs(5126)
crs5126 # look at the whole CRS
crs5126$Name # pull out just the name of the crs
#[1] "WGS 84."
crs5126$wkt # crs in well-known text format
```


# Feature collection (points)

```{r feature collection}
sdata
```


# Cartography

```{r cartography}
plot(sdata["richness_raw"])

plot(sdata)
```


# Thematic maps for communication

```{r Thematic maps for communication}
#using tmap

tm1 <- tm_shape(sdata) + 
  tm_dots(col = "richness_raw", 
          palette = "plasma", 
          size = 0.5, 
          title = "Copepod richness")
tm1

tmap_save(tm1, filename = "../code/outputs/figures/Richness-map.png", 
          width = 600, height = 600)
```


# Mapping spatial polygons as layers

## Loading shapefiles

```{r load shapefiles}
aus <- st_read("../data/workshop4/spatial-data/Aussie/Aussie.shp")
aus

shelf <- st_read("../data/workshop4/spatial-data/aus_shelf/aus_shelf.shp")
shelf
```

## Mapping your polygons

```{r map polygons}
tm_shape(shelf) + 
  tm_polygons()

tm_shape(shelf, bbox = sdata) + 
  tm_polygons() +
  tm_shape(aus) + 
  tm_polygons() + 
  tm_shape(sdata) + 
  tm_dots()
  
# make sure that aus, shelf and sdata are all in the same CRS before plotting together
st_crs(aus)
st_crs(shelf)
st_crs(sdata)
# transform sdata to match the others
sdata <- st_transform(sdata, st_crs(shelf))
```


# Exploring t_map

```{r exploring tmap}
?tmap

tmap_style("beaver")
```


```{r tmap vignette}
vignette('tmap-getstarted')
```




