---
title: "R for Marine Science Workshop 1"
author: "Darius HUISMAN"
date: "2026-06-03"
output: 
 html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 6
    fig_width: 10
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: inline
---


```{r setup, include=FALSE}
# Set options, housekeeping
knitr::opts_chunk$set(
	echo = FALSE,
	include = TRUE,
	message = FALSE,
	warning = FALSE)
```

## Introduction


### Clear out environment
```{r housekeeping, message=FALSE, warning=FALSE, include=FALSE}
rm(list=ls())
```

### Load packages
```{r packages, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
```

### Source script
```{r source-script1, message=FALSE, warning=FALSE, include=FALSE}
source("Workshop1.R")
```


## 1.9 Integrating data grammar with visual diagnostics in qmd

###1.9.1 Wrangling and Plotting in Parallel

```{r Wrangling and Plotting in Parallel}
#summary table
penguins  |>
  group_by(species, island) |>
  summarise(mean_bodymass = mean(body_mass_g, na.rm = TRUE)) 
penguins|> 
  ggplot(aes(x = species, y = body_mass_g, fill =island)) + 
  geom_boxplot()  
```

###1.9.2 Piping Directly to Visualisation
```{r Piping Directly to Visualisation}
# Pipe directly from aggregation to plotting with error bars
mass_compare_plot <- penguins |>
  group_by(species, island) |>
  summarise(
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    sd_mass = sd(body_mass_g, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |>
  ggplot(aes(x = species, y = mean_mass, colour = island)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_mass - sd_mass, 
                    ymax = mean_mass + sd_mass), 
                width = 0.2) +
  labs(title = "Mean Body Mass by Species and Island",
       subtitle = "Error bars represent standard deviation",
       y = "Mean Body Mass (g)",
       x = "Species") +
  theme_minimal()

mass_compare_plot

```

Pause for thought: Think about what this figure is telling you about the biology of these species of penguins on these different islands. What do you think a statistical test would tell you here?
  The bodymass correlates with the species and not the island.




```{r challenge (SE instead of SD)}
# Pipe directly from aggregation to plotting with error bars
mass_compare_plot <- penguins |>
  group_by(species, island) |>
  summarise(
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    se_mass = (sd(body_mass_g, na.rm = TRUE)/sqrt(length(body_mass_g))),
    n = n(),
    .groups = "drop"
  ) |>
  ggplot(aes(x = species, y = mean_mass, colour = island)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_mass - se_mass, 
                    ymax = mean_mass + se_mass), 
                width = 0.2) +
  labs(title = "Mean Body Mass by Species and Island",
       subtitle = "Error bars represent standard errors",
       y = "Mean Body Mass (g)",
       x = "Species") +
  theme_minimal()

mass_compare_plot

```


##1.10 Saving, exporting, and version milestones

```{r Create output directories if they do not exist}
# Create output directories if they do not exist:
if (!dir.exists("outputs/figures")) dir.create("outputs/figures") # folder for figs
if (!dir.exists("outputs/tables")) dir.create("outputs/tables") # folder for tables
if (!dir.exists("Rdata")) dir.create("Rdata") # folder for Rdata objects

```

```{r export tables}
# 1. Exporting our collapsed summary table as a universal flat text file
write_csv(biological_signal, "outputs/tables/penguin_species_mass_summary.csv")

# 2. Saving our penguins_final table as a native R binary file
saveRDS(penguins_final, "outputs/tables/penguins_final.rds")

```

```{r exprot figures}
ggsave("outputs/figures/mass_compare_plot.png", 
       plot = mass_compare_plot, 
       width = 120, height = 120, 
       units = "mm", dpi = 300)

```

